Concentration of Maxima and Fundamental Limits in High-Dimensional Testing and Inference

### 2021-04-19 reviews/response.tex

%    and the signal-size parameter of the corresponding chi-square association test.  This allows us to quantify the statistical power and
%     optimal design questions in a unified manner as well as ultimately explain the role of phase-transitions in the fundamental statistical limits of GWAS.
%

  \item {\bf (Organization)} 
  The manuscript has been revised substantially and now includes a brief introductory chapter and 6 substantive chapters, which amount to a total of 125 pages.
  As with the original submission, Chapter 2 reviews the statistical risks, estimation and testing procedures, as well as probability models for the errors used in the rest of the manuscript.

  After very careful deliberation, we decided to keep a majority of the content in Chapter 3, which provides a panorama of the phase-transition results in the signal detection and support estimation problems for the additive iid Gaussian error model.
  There were some mixed opinions from the reviewers regarding the inclusion of a chapter that paints the landscape in broad strokes and provides references to the state-of-the-art of the literature, at the expense of delaying entirely original contributions.
  We think that providing an accessible account of the subject of high-dimensional multiple testing would benefit readers who are not already experts in the area, and might just narrowly outweigh the additional patience required on the part of experts before we get to the more exciting material.
  Details and explanations of the changes we made can be found in the reponse to Reviewer 2 (R2) below.

@@

  To better balance the exposition, we reorganized Chapter 4 of the previous version of the draft into two new chapters (now Ch.\ 4, entitled ``Exact Support Recovery Under Dependence'',  and Ch.\ 5 ``Bayes and Minimax Optimality''). 
  This change, in addition to making all chapters more manageable and of about equal lengths, exposes more clearly our contributions. 
  The results contained in Chapters 4 and 5 as well as the more probabilistic Chapter 6 and the more applied Chapter 7 are original contributions to the literature.

  Following the suggestion of R2,  we have placed most proofs in the text after the corresponding claims (theorems, propositions, lemmas).  In the interest of space and to conform with the format of the SpringerBriefs series, we no longer include ``Exercises'' sections.

  \item {\bf (Typos, grammar, and style)}  Prompted by the suggestions of R2, we have revised the language of the manuscript. 
  There have been extensive re-writes throughout the manuscript in order to make the content more connected and style more consistent.  
  In trying to make the text more accessible, we re-formatted the statements of most theorems to make them more modular. 
  We have also corrected typos and missing articles.

@@

    \medskip
    Thank you.  We were not aware of the interesting and very recent work \cite{li2020optimality} of Li and Fithian when the first draft was prepared. We wholeheartedly agree that the set-up described by the reviewer is much more realistic.
    It is stimulating for us to learn that the Higher Criticism statistic no longer dominates the max-type Bonferonni thresholding procedure when $1/2<\beta<3/4$, as long as the signals are spread out, even if the noise terms remain Gaussian.
%    This changed our understanding about optimality of the max-statistic.  
    We now discuss and reference this recent and fascinating development in Chapters 2 and 3 where appropriate. 
%    This new development on the spread-out and heavy-tailed signals setting is parallel to the heavy-tailed noise setting that we pursued in this manuscript.

    One naturally wonders what happens, e.g., when both signals and noise terms are distributed with regularly-varying tails, and what can be said about the support recovery problems when signals are spread-out.
    Since our focus in this manuscript is weighted towards support estimation problems rather than the detection problem, we have refrained from discussing more in the direction of the more realistic setting of \cite{li2020optimality}, which would require resolving some new mathematical challenges. 
    We hope to understand this phenomenon better in the near future.

    \item {\em The statistical risks for exact-approximate and approximate-exact support recovery look artificial and uninterpretable. 
    Could you provide further explanations of these two criteria?}

    We have edited the following explanation when introducing the asymmetric statistical risks:
    \begin{quote}
    $\dots$
	In applications, however, attitudes towards false discoveries and missed signals are often asymmetric.
	In the example of GWAS, where the number of candidate locations $p$ could be in the millions, and a class imbalance between the number of nulls and signals exists, researchers are typically interested in the marginal (location-wise) power of discovery, while exercising stringent (family-wise) false discovery control. 
	These types of asymmetric considerations, while important in applications, have not been studied theoretically.  
	For example, the GWAS application motivates the \emph{exact-approximate} support recovery risk, which weighs both the family-wise error rate and the marginal power of discovery:
	$\dots$
    \end{quote}
    In fact, we believe that power calculations for GWAS are most natrually interpreted when viewed through the exact-approximate support recovery risk.

@@

    \item {\em The uniform relative stability (URS) is type of weak dependence based on the results in Chapter 5. In FDR literature, the statistics are 
    said to be weakly dependent if 
    $$
    \frac{1}{p^2} \sum_{i=1}^n |{\rm Cor}(x_i,x_j)| = o(1),
    $$
    see e.g. Efron 2007 and Fan 2012. Is there any connection between these two types of weak dependence?}

    It can be shown that the more restrictive UDD' condition (Definition 6.2 in the new numbering) implies weak dependence as described in Fan et al \cite{fan:xu:gu:2012} (the single sum in the reviewer's comment seems to be a typo),
    $$
    \lim_{p\to\infty}\frac{1}{p^2} \sum_{i,j} |\Sigma_p(i,j)| = 0.
    $$
    Starting with the assumed permutations $l_p$ of $\{1, . . . , p\}$ in UDD', we have the equality
    $$
    \sum_{i,j} |\Sigma_p(i,j)| = \sum_{i,j} |\Sigma_p(i',j')|,
    $$
    where $i'= l_p(i)$, $j'= l_p(j)$, since permutations do not affect the sum.
    By the second assumption of a sequence $r_n\to0$ that bounds the correlations from above, we have
    $$
    \frac{1}{p^2} \sum_{i,j} |\Sigma_p(i',j')| \le \frac{1}{p^2} \sum_{i,j} r_{|i-j|} \le \frac{1}{p} \sum_{k=0}^{p-1} r_{k} \to 0.
    $$
    Therefore UDD' impies the weak dependence condition.

    \medskip
    We can also establish that the UDD/URS condition is no-weaker than the weak dependence condition.
    We use the same non-UDD example outlined in Example 6.1 (new numbering).
    Suppose ${\cal E} = \left(\epsilon_p(i)\right)_{i=1}^p$ is Gaussian, and is comprised of $\lfloor p^{1-\beta}\rfloor$ blocks, each of size at least $\lfloor p^\beta \rfloor$.  Let the elements within each block have correlation 1, and let the elements from different blocks be independent. 
    It follows that 
    $$
    \frac{1}{p^2} \sum_{i,j} |\Sigma_p(i,j)| \sim \frac{1}{p^2} (p^{1-\beta})^2 p^{\beta} = p^{-\beta} \to 0,
    $$
    and therefore ${\cal E}$ is weakly dependent as defined in \cite{fan:xu:gu:2012}, but not UDD/URS.

    However, it is not clear to us whether the weak dependence condition is always weaker than UDD.
    The comparison is complicated by the fact that the definition of UDD only cares about positive correlations, while the weak dependence condition takes the sum of absolute values of the correlations.

@@

%  accustomed to deal with in the case of sums/averages.
%

    \item {\em This is another line of work that studies the strong dependence structure where the covariance matrix has a factor structure (eg. Fan 2012). For instance, this includes the equi-correlated case. Are you aware of any results under strong dependence in your context?}

    We thank the reviewer for pointing to the interesting methodological work of Fan, Xu and Gu \cite{fan:xu:gu:2012}, which aimed to estimate false discovery proportions for heavily dependent observations. 
    We are not aware of works that discuss optimality or feasibility of support recovery (i.e., multiple testing) problems under such dependence conditions.
    The results in our text in their current form do not easily extend to that of Fan et al., since even the remainder term $K_i$'s in the factor model can be more heavily dependent than under the UDD/URS condition, as discussed in the response to the previous question.
    Nevertheless, Fan et al provides a glimpse into how heavy dependence (e.g., the equi-correlated case) may be handled, and how theoretical analysis may be carried out. We do hope to develop a theory that applies to the more general setting.
%   It can be seen that the {\em weak dependence} condition in their Definition 1, implies our UDD condition (for a correlation matrix).  
%   Therefore, the Gaussian error arrays of the $K_i$-variables in Relation (10) in the latter reference, are URS (uniformly relatively stable).  
%   This suggests that the PFA methodology in \cite{fan:xu:gu:2012} can perhaps be extended to apply to the exact support recovery problem.  
%   That is, one can project on the orthogonal complement of the space spanned by the top$-k$ eigenvectors of the error-covariance.  Then, perhaps 
%   under standard in-coherence conditions, between the  signal and the covariance, the simple Bonferonni thresholding estimator is optimal for exact 
%   support recovery.  
%   We are not aware of many more examples, but hope that the developed general theory applies in many more setting.

@@

    \item {\em Proposition 4.3 still requires $f_a$ to be MLR, right? If so, please state it.}

    Thank you for the sharp observation.  The previous presentation was not sufficiently clear.  We have now devoted the new Ch.\ 5 to the
    finite-sample Bayes optimality (and minimax optimality) to better present these topics.  
    The former Proposition 4.3 appears now as Theorem 5.2.  The MLR property on $f_a$
    is no longer necessary.  This is because the {\em likelihood ratio theresholding} estimator is defined in such a way (in terms of the likelihood) that
    the desired monotonicity property is guaranteed.  The MLR property was essential in Theorem 5.1 since there we deal with thresholding (as opposed to
    likelihood thresholding). The MLR property basically ensures that in the setting of Theorem 5.1, the simple thresholding estimators are in fact of
    {\em likelihood thresholding} type.  We realize that Theorem 5.2 is a bit too theoretical (i.e., not directly useful in practice since we don't have the 
    likelihood) but it provides some insights on
    the general Bayes (sub)optimal support estimation procedures.

    \item  {\em Fig 3.2 shows that the exact-approximate support recovery is strictly easier to achieve than the approximate-exact support recovery. Do you have intuition on it?}

    To provide some intuition, consider the distances between the decision boundaries and the means of the null and signal part of the observations, respectively.
    Controlling the FWER requires setting the decision boundaries approximately 
    \begin{equation} \label{FWER}
    \sqrt{2\log{p}}
    \end{equation}
    away from zero (mean of nulls), as is well-known in the Bonferroni adjustment.

    Controlling FDR, on the other hand, depends on the number of signals present.
    When there are fewer signals, a FDR-controlling procedure has to be more conservative -- when the number of true signals is small, the number of false discoveries reported along with them has to reduce as well.
    Consequently, the decision boundary has to be further away from zero as $\beta\to1$.
    Quantitatively, this distance varies with $\beta$ approximately as
    \begin{equation} \label{FDR}
    \sqrt{2\beta\log{p}},
    \end{equation}
    with sparser problems requiring a larger separation. 
    Notice that the distance required for FDR control in an extremely sparse problem ($\beta\to1$) is no different from that of FWER control.
    For very dense problems ($\beta\to0$), the distance vanishes on the Bonferroni scale.

@@

    We have clarified the terminology througout.  In particular, Chapter 2 is a now entitled ``Risks, Procedures, and Error Models''.

    \end{enumerate}

We thank the reviewer for pointing a number of typos, which we have corrected.  We respond below to a subset of the minor points
and comments raised in the report.

\medskip
\noindent {\bf Typos and other minor comments}
\begin{itemize}
	\item [8.] {Page 25, what does the dark-gray area that is below the area marked by
	``M'' and on the right of the area marked by ``$S, L_1 , L_2$'' correspond to?}

    The area is detectable by the HC statistic. We have made changes to the figure and clarified this in the caption.

	\item [9.] {Chapter 3, the notation $f(\beta), g(\beta), h(\beta), \widetilde{g}(\beta), \widetilde{h}(\beta)$ is hard to memorize. What about $f_P(\beta)$ where $P\in\{\mathrm{D, A, E, AE, EA}\}$?}

	Agreed! We have made changes everywhere in the text.

	 \item [10.] {\em Page 46, ``these Hamming loss-minimax studies naturally amounts to the analysis of the elementary case of iid data,...'' 
	 Is this true? Although the indicators can be dealt with separately but the threshold may depend on the entire dataset as in Holm, 
	 Hochberg, and BH procedures. }

	% You are right that our statement was a bit exaggerated and thus perhaps misleading.
	Thank you for the feedback. We have clarified and tempered the language:
	\begin{quote}
	 More importantly, since Hamming loss decomposes into expectations on individual terms that are not affected by dependence, Hamming loss-minimax studies do not reveal the difference in probability of support recovery between independent and dependent observations. 
	\end{quote}
	Our analysis, and e.g. that of Butucea et al \citep{butucea2018variable}, consider all thresholding procedures which include data-dependent procedures that are affected by dependence.

@@

    Thank you for the insightful comment.  Indeed, by Slepian's Lemma, negatively correlated Gaussian variables have larger maxima (in stochastic order)
    than iid maxima (with the same marginals).  This difference, however, is negligible asymptotically.  Intuitively, this is due to the fact that while Gaussian vectors can be equi-correlated with correlation $\rho$ that can be arbitrarily close to $1$, this common correlation is bounded below by $-1/(p-1)$, where $p$ is the number of the variables.

    Our choice of words was imprecise. We should have said that the rate of growth of dependent maxima is no faster than in the iid case.  
    We now clarify this point in the new Remark 2.1 (see Relation (2.41)).

    This general observation is entirely due to the light-tailed (rapidly varying) nature of the marginal distributions.  It applies to error-arrays with arbitrary
    dependence and the proof involves the union bound and the rapid variation property of the marginal distributions.  Please see Proposition 2.2.

    \item[12.] {Page 52, scenario 1, $r^{(2)} = (1 + \delta) \implies r^{(2)} = (1 + \delta)g(\beta)$.}

    Thank you for your careful reading of the text! 
    However this was not a typo -- we wanted to provide an example where signals sizes straddle the boundary but exact support recovery is still achieved. 
    There is only one signal of size $r^{(2)}$ which corresponds to the required signal size at its sparsity $\beta=1$, i.e., $r^{(2)} = (1 + \delta)g(1)$.
%    We have changed the expression to
%    $$r^{(2)} = (1+\delta)f_{\mathrm{E}}(1)$$
%    for clarification.

### 2021-04-19 2.background/2.statistical_risks.tex
updated Ch 4
I updated up to and including Ch 4.2.2.

Hope to finish tomorrow.

{\bf Signal detection.} 
Recall that in sparse signal detection problems, our goal is to come up with a procedure, $\mathcal{R}(x)$, such that the null hypothesis is rejected if the data $x$ is deemed incompatible with the null.
In the additive error models context \eqref{eq:model-additive}, we wish to tell apart two hypotheses
\begin{equation} \label{eq:global-test-additive}
    \mathcal{H}_0: \mu(i) = 0, \;i=1,\ldots,p,
    \quad\text{v.s.}\quad 
    \mathcal{H}_1: \mu(i)\neq 0, \; \text{for some }i\in\{1,\ldots,p\},
\end{equation}
based on the $p$-dimensional observation $x$.
Similarly, in the chi-square model \eqref{eq:model-chisq}, we look to test
\begin{equation} \label{eq:global-test-chisq}
    \mathcal{H}_0: \lambda(i) = 0, \;i=1,\ldots,p,
    \quad\text{v.s.}\quad 
    \mathcal{H}_1: \lambda(i)\neq 0, \; \text{for some }i\in\{1,\ldots,p\}.
\end{equation} 
Since the decision is binary, we may write the outcome of the procedure in the form of an indicator function, 
$T(\mathcal{R}(x))\in\{0,1\}$, where $T=1$ if the null is to be rejected in favor of the alternative, and 0 if we fail to reject the null. 
The Type I and Type II errors of the procedure, i.e., the probability of wrong decisions under the null hypothesis $\mathcal{H}_0$ and alternative hypothesis $\mathcal{H}_1$, respectively, are defined as 
\begin{equation} \label{eq:type-II-error}
    \alpha(\mathcal{R}) := \P_{\mathcal{H}_0}\left(T()\mathcal{R}(x))=1\right)
    \quad \text{and} \quad
    \beta(\mathcal{R}) := \P_{\mathcal{H}_1}\left(T(\mathcal{R}(x))=0\right).
\end{equation}
The Neyman-Pearson framework of hypothesis testing then seeks tests that minimize the Type II error of the test, while controlling the Type I error of the test at low levels.
We are particularly interested in the sum of the two errors, 
\begin{equation} \label{eq:risk-detection}
    \mathrm{risk}^{\mathrm{D}}(\mathcal{R}) := \alpha(\mathcal{R}) + \beta(\mathcal{R}),
\end{equation}
which shall be referred to as the risk of signal detection (of the procedure $\mathcal{R}$).
It is trivial that a small $\mathrm{risk}^{\mathrm{D}}$ would imply both small Type I and Type II errors of the procedure.

@@

Notice that all risk metrics introduced so far penalize false discoveries and missed signals somewhat symmetrically --- the approximate support recovery risk combines proportions of errors, the exact support recovery risk combines probabilities of errors, and the Hamming loss increments the risk by one regardless of the types of errors made.
In applications, however, attitudes towards Type I and Type II errors are often different.
In the example of GWAS, where the number of candidate locations $p$ could be in the millions, researchers are typically interested in the marginal (location-wise) power of discovery, while exercising stringent (family-wise) false discovery control. 
Such types of asymmetric risks, while important in applications, have not been studied theoretically.  
This asymmetric consideration, and in particular the GWAS application, prompts us to consider risks that weigh both the family-wise error rate and the marginal power of discovery.
One such risk metric is what shall be referred to as the \emph{exact-approximate} support recovery risk
\begin{equation} \label{eq:risk-exact-approx}
    \mathrm{risk}^{\mathrm{EA}}(\mathcal{R}) := \mathrm{FWER}(\mathcal{R}) + \mathrm{FNR}(\mathcal{R}).
\end{equation}
The somewhat cumbersome name and notation are chosen to reflect
the asymmetry in dealing with the two types of errors in support recovery.
Namely, when the risk metric \eqref{eq:risk-exact-approx} vanishes, we have ``exact false discovery control, and approximate false non-discovery control'' asymptotically.

>>>>

{\bf Signal detection.} 
Recall that in sparse signal detection problems, our goal is to come up with a procedure, $\mathcal{R}(x)$, such that the null hypothesis is rejected if the data $x$ is deemed incompatible with the null.
In the additive error models context \eqref{eq:model-additive}, we wish to tell apart two hypotheses
\begin{equation} \label{eq:global-test-additive}
    \mathcal{H}_0: \mu(i) = 0, \;i=1,\ldots,p,
    \quad\text{v.s.}\quad 
    \mathcal{H}_1: \mu(i)\neq 0, \; \text{for some }i\in\{1,\ldots,p\},
\end{equation}
based on the $p$-dimensional observation $x$.
Similarly, in the chi-square model \eqref{eq:model-chisq}, we look to test
\begin{equation} \label{eq:global-test-chisq}
    \mathcal{H}_0: \lambda(i) = 0, \;i=1,\ldots,p,
    \quad\text{v.s.}\quad 
    \mathcal{H}_1: \lambda(i)\neq 0, \; \text{for some }i\in\{1,\ldots,p\}.
\end{equation} 
Since the decision is binary, we may write the outcome of the procedure in the form of an indicator function, 
$T(\mathcal{R}(x))\in\{0,1\}$, where $T=1$ if the null is to be rejected in favor of the alternative, and 0 if we fail to reject the null. 
The Type I and Type II errors of the procedure, i.e., the probability of wrong decisions under the null hypothesis $\mathcal{H}_0$ and alternative hypothesis $\mathcal{H}_1$, respectively, are defined as 
\begin{equation} \label{eq:type-II-error}
    \alpha(\mathcal{R}) := \P_{\mathcal{H}_0}\left(T(\mathcal{R}(x))=1\right)
    \quad \text{and} \quad
    \beta(\mathcal{R}) := \P_{\mathcal{H}_1}\left(T(\mathcal{R}(x))=0\right).
\end{equation}
The Neyman-Pearson framework of hypothesis testing then seeks tests that minimize the Type II error of the test, while controlling the Type I error of the test at low levels.
We are particularly interested in the sum of the two errors, 
\begin{equation} \label{eq:risk-detection}
    \mathrm{risk}^{\mathrm{D}}(\mathcal{R}) := \alpha(\mathcal{R}) + \beta(\mathcal{R}),
\end{equation}
which shall be referred to as the risk of signal detection (of the procedure $\mathcal{R}$).
It is trivial that a small $\mathrm{risk}^{\mathrm{D}}$ would imply both small Type I and Type II errors of the procedure.

@@

Notice that all risk metrics introduced so far penalize false discoveries and missed signals somewhat symmetrically --- the approximate support recovery risk combines proportions of errors, the exact support recovery risk combines probabilities of errors, and the Hamming loss increments the risk by one regardless of the types of errors made.
In applications, however, attitudes towards false discoveries and missed signals are often asymmetric.
In the example of GWAS, where the number of candidate locations $p$ could be in the millions, and a class imbalance between the number of nulls and signals exists, researchers are typically interested in the marginal (location-wise) power of discovery, while exercising stringent (family-wise) false discovery control. 
These types of asymmetric considerations, while important in applications, have not been studied theoretically.  
For example, the GWAS application motivates the \emph{exact-approximate} support recovery risk, which weighs both the family-wise error rate and the marginal power of discovery:
\begin{equation} \label{eq:risk-exact-approx}
    \mathrm{risk}^{\mathrm{EA}}(\mathcal{R}) := \mathrm{FWER}(\mathcal{R}) + \mathrm{FNR}(\mathcal{R}).
\end{equation}
The somewhat cumbersome name and notation are chosen to reflect
the asymmetry in dealing with the two types of errors in support recovery.
Namely, when the risk metric \eqref{eq:risk-exact-approx} vanishes, we have ``exact false discovery control, and approximate false non-discovery control'' asymptotically.
